---
title: "APOE Isoform Exploratory Analysis — Colorectal Cancer (TCGA)"
author: "Ghani"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_width: 8
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center",
  cache      = TRUE
)

# Load configuration
library(yaml)
params <- read_yaml("../config/params.yaml")
```

# 1. Overview

This notebook provides an interactive exploratory analysis of APOE isoform
distributions and their associations with clinical, genomic, and immune
features in TCGA colorectal cancer (COAD/READ) cohorts.

---

# 2. Load Libraries

```{r libraries}
library(tidyverse)
library(DESeq2)
library(survival)
library(survminer)
library(ComplexHeatmap)
library(ggpubr)
```

# 3. Load Data

```{r load-data}
# Clinical + isoform data
clinical <- read_csv(file.path(params$paths$processed_data, "clinical_with_isoforms.csv"))

# Expression matrix (normalized)
expr <- read_csv(file.path(params$paths$processed_data, "expression_normalized.csv"))

# Immune infiltration scores
immune <- read_csv(file.path(params$paths$processed_data, "immune_infiltration.csv"))

# Mutation data
mutations <- read_csv(file.path(params$paths$processed_data, "mutation_summary.csv"))
```

---

# 4. APOE Isoform Distribution

```{r isoform-distribution, fig.cap="APOE isoform frequencies in TCGA CRC cohort"}
clinical %>%
  count(apoe_isoform) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = reorder(apoe_isoform, -n), y = n, fill = apoe_isoform)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", pct)), vjust = -0.5) +
  scale_fill_manual(values = c(
    "e2" = params$figures$color_palette$e2,
    "e3" = params$figures$color_palette$e3,
    "e4" = params$figures$color_palette$e4
  )) +
  labs(
    title = "APOE Isoform Frequencies in CRC (TCGA)",
    x = "APOE Isoform Group",
    y = "Number of Patients"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

---

# 5. Demographics by Isoform

```{r demographics}
clinical %>%
  group_by(apoe_isoform) %>%
  summarise(
    n           = n(),
    age_mean    = round(mean(age_at_diagnosis, na.rm = TRUE), 1),
    age_sd      = round(sd(age_at_diagnosis, na.rm = TRUE), 1),
    pct_female  = round(mean(gender == "FEMALE", na.rm = TRUE) * 100, 1),
    .groups     = "drop"
  ) %>%
  knitr::kable(caption = "Patient demographics by APOE isoform group")
```

---

# 6. APOE Expression by Isoform

```{r apoe-expression, fig.cap="APOE mRNA expression across isoform groups"}
# Extract APOE expression and merge with clinical
apoe_expr <- expr %>%
  filter(gene == "APOE") %>%
  pivot_longer(-gene, names_to = "sample_id", values_to = "expression") %>%
  left_join(clinical, by = "sample_id")

ggboxplot(
  apoe_expr,
  x      = "apoe_isoform",
  y      = "expression",
  fill   = "apoe_isoform",
  palette = c(
    params$figures$color_palette$e2,
    params$figures$color_palette$e3,
    params$figures$color_palette$e4
  ),
  add    = "jitter",
  add.params = list(alpha = 0.3)
) +
  stat_compare_means(method = "kruskal.test", label.y.npc = 0.95) +
  labs(
    title = "APOE mRNA Expression by Isoform Group",
    y     = "Normalized Expression (log2)"
  ) +
  theme(legend.position = "none")
```

---

# 7. Immune Infiltration Overview

```{r immune-heatmap, fig.cap="Immune cell infiltration by APOE isoform", fig.height=8}
# Prepare matrix for heatmap
immune_matrix <- immune %>%
  left_join(clinical %>% select(sample_id, apoe_isoform), by = "sample_id") %>%
  group_by(apoe_isoform) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = "drop") %>%
  column_to_rownames("apoe_isoform") %>%
  as.matrix() %>%
  t() %>%
  scale()

Heatmap(
  immune_matrix,
  name = "Z-score",
  cluster_columns = FALSE,
  row_names_gp    = gpar(fontsize = 9),
  column_title    = "Mean Immune Cell Infiltration by APOE Isoform"
)
```

---

# 8. Survival — Kaplan-Meier (OS)

```{r km-os, fig.cap="Overall Survival by APOE Isoform"}
fit_os <- survfit(
  Surv(OS.time, OS) ~ apoe_isoform,
  data = clinical
)

ggsurvplot(
  fit_os,
  data       = clinical,
  pval       = TRUE,
  risk.table = TRUE,
  palette    = c(
    params$figures$color_palette$e2,
    params$figures$color_palette$e3,
    params$figures$color_palette$e4
  ),
  xlab       = "Time (days)",
  ylab       = "Overall Survival Probability",
  title      = "OS by APOE Isoform in CRC"
)
```

---

# 9. Quick Summary Statistics

```{r summary-stats}
list(
  total_patients     = nrow(clinical),
  isoform_counts     = table(clinical$apoe_isoform),
  median_follow_up   = median(clinical$OS.time, na.rm = TRUE),
  overall_event_rate = mean(clinical$OS, na.rm = TRUE)
) %>%
  str()
```

---

# 10. Session Info

```{r session-info}
sessionInfo()
```
